---
title: "Flattening lists"
description: |
  In this post, I explore various ways of flattening hierarchical lists in 
  Python, contrasting with the built-in `unlist` function in R
author:
  - name: Abhijit Dasgupta
    url: {}
date: 2022-02-23
output:
  distill::distill_article:
    self_contained: false
    toc: true
    toc_float: true
draft: true
categories:
  - R
  - Python
  - data munging
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introducing the problem

Today, we will focus on the results of splitting text fields by a delimiter
in R and Python. This will typically result in a **list of lists**. However, 
that would not do for my purpose. 

The situation I was facing was to parse a text field containing one or more comma-separated disease names that were afflicting each patient in a study, and I wanted
to make a frequency distribution of all the diseases among patients in the study, knowing
that each disease would only appear at most once per patient. So I had to **flatten**
the list of lists I would get by parsing the text field, so that I could then 
send it to a tabulation function like `table` in R or `numpy.unique(..., return_counts=True)` in Python.

# First attempt

Let's start with a text array containing the data in question.This data is an
excerpt from a dataset provided for the [PKDD '99 Discovery Challenge](https://sorry.vse.cz/~berka/challenge/pkdd1999/tsumoto.htm)

```{r}
dat = readr::read_csv('dat.csv')
head(dat)
```

There are comma-separated diseases here. We'll ignore the obvious need to clean
the text field of extraneous spaces for now. Let's try separating out the different
diseases.

```{r}
stringr::str_split(dat$Diagnosis, ",")[1:5]
```

We see that this is a list of vectors (so we can nominally consider it a list of lists), 
some of length 1 and some with length more than 1. In order to tabulate the data,
all we need is the `unlist` function.

```{r}
table(unlist(stringr::str_split(dat$Diagnosis, ",")))
```

#### Voila!!!

# Doing this in Python

It turns out, doing a similar operation would also get you a list of lists
in Python

```{python}
import pandas as pd
dat = pd.read_csv('dat.csv')
d = dat.Diagnosis.str.split(",")
d[:5]
```
It turns out, we don't have an `unlist` function in Python!!

## Finding a resource
Google being the all-encompassing deity of knowledge, I turned there and 
found [this post](https://www.pythonpool.com/flatten-list-python/) describing 
25 ways to flatten a list (and obviously a Simon and Garfunkel fan).
There are a couple of one-liners in that post that I'll highlight here.

## The common method

Searches for this problem most commonly provide the following solution, either 
as a for-loop or a list comprehension

```{python}
flatten_list = []
for sublist in d:
	for item in sublist:
		flatten_list.append(item)
flatten_list[:8]
```

## Functional programming and `reduce`

```{python}
from functools import reduce
reduce(list.__add__, d)[:8]
```
Functional programming provides functions _that act on functions_. Here, the 
`reduce` function applies the `list.__add__` function recursively to successive
elements of `d`, noting that _adding_ two lists serves to concatenate them in 
Python.

A variant on this theme that works is 

```{python}
sum(d, [])[:8]
```
since the `sum` function applies addition recursively from a starting value, which is
the second argument. Here the addition operation is overloaded for lists to 
concatenate lists.

## Using `pandas`

**pandas** does have a `flatten` function hidden in its depths

```{python}
from pandas.core.common import flatten
list(flatten(d))[:8]
```

## Using `itertools`

```{python}
import itertools
list(itertools.chain(*d))[:8]
```




