---
title: "Tooltips using ggplotly"
description: |
  In this post I highlight how to format tooltips from the ggplotly interface
  to plotly from R
author:
  - name: Abhijit Dasgupta
    url: {}
date: 2022-03-01
output:
  distill::distill_article:
    self_contained: false
categories:
  - r
  - data visualization
---

```{css, echo=FALSE}
@import url('https://fonts.googleapis.com/css2?family=Roboto+Slab:wght@500&display=swap');
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, messages=TRUE, warning = FALSE)
library(ggplot2)
theme_set(theme_bw() + theme(text = element_text(family='Roboto Slab')))
```

You can incorporate and format tooltips in the `ggplotly` interface. The `ggplotly`
function has a `tooltip` option that can be inhabited by aesthetics in the 
`ggplot` code. By default it shows all the aesthetics specified in the ggplot code. 

In the following plot, all of the aesthetics (`bill_length_mm`, `body_mass_g` and `species`) are displayed in the tooltips.

```{r}
pacman::p_load(plotly, ggplot2, palmerpenguins)
plt <- ggplot(penguins, aes(x = bill_length_mm, y = body_mass_g))+
	geom_point(aes(color = species)) + 
	labs(x = 'Bill length (mm)', y = 'Body mass (g)', 
		 title = 'Penguins')

ggplotly(plt)
```

You can specify particular variables to display in tooltips using the `tooltips` 
option in `ggplotly`

```{r}
pacman::p_load(plotly, ggplot2, palmerpenguins)
plt <- ggplot(penguins, aes(x = bill_length_mm, y = body_mass_g))+
	geom_point(aes(color = species)) + 
	labs(x = 'Bill length (mm)', y = 'Body mass (g)', 
		 title = 'Penguins')

ggplotly(plt, tooltip = c('x','y'))

```

One can trick `ggplotly` to format a tooltip, using the `text` aesthetic. You can use
this in geometries where `text` doesn't have an effect, and you can format this
aesthetic using R methods, like `glue::glue` and `format`

```{r, warning=TRUE}
pacman::p_load(plotly, ggplot2, palmerpenguins)
plt <- ggplot(penguins, aes(x = bill_length_mm, y = body_mass_g))+
	geom_point(aes(
	  color = species, 
		text = glue::glue("<b>{species}</b><br>Body mass: {body_mass_g} g<br>Bill length {bill_length_mm} mm"))) + 
	labs(x = 'Bill length (mm)', y = 'Body mass (g)', 
		 title = 'Penguins')

ggplotly(plt, tooltip = c('text'))

```

> If you just try to run the ggplot2 code, you'll get the warning "Ignoring unknown
aesthetics: text". We're taking advantage of this to allow it to not show in the plot
but provide tooltip data to `ggplotly`


